\section{Introduction}

Jahob is a system for verification of data structure
consistency.  It builds on our previous system Hob
\cite{LamETAL05GeneralizedTypestateCheckingDataStructureConsistency,
LamETAL04HobProjectWebPage,
LamETAL04OurExperienceModularPluggableAnalyses,
LamETAL03OnModularPluggableAnalysesUsingSetInterfaces,
LamETAL05HobTool,
LamETAL05CrossCuttingTechniquesProgramSpecificationAnalysis} and
generalizes Hob to deal with dynamic instantiation of individual data
structures and a richer specification language for procedure
contracts, which allows describing properties of in terms of not only
sets but also relations.  It is related to projects such as ESC/Java
\cite{FlanaganETAL02ExtendedStaticCheckingJava}, ESC/Java2
\cite{CokKiniry04ESCJava2} and Spec\#
\cite{BarnettETAL04SpecProgrammingSystemOverview}, but has
more emphasis on data structure consistency and automation
through a combination of heterogeneous static analyses.

\subsection{Design Decisions}

We next motivate several design decisions in Jahob.
\begin{itemize}

\item Jahob uses a small subset of Java as an implementation language
(see Section~\ref{sec:javaSubset}).

\item Jahob uses a subset of Isabelle formulas as the specification language and as
a communication language between different analysis plugins and
decision procedures.

\item Jahob views classes as the main units of encapsulation, not packages, and not objects.
The fact that objects are not the unit of encapsulation means that the
notions of an invariant, abstract variable, and private state are all
defined on the collection of all objects of the class (along with
potential \emph{claimed} fields from other classes).  This view, where
object fields are global binary relations and encapsulation can be
achieved using simple syntactic mechanisms, makes soundness arguments
simple and supports data structure implementations that involve
collaborations between different objects of the class.  Properties
that involve defining per-object state of instantiatable data
structures become a consequence of the global simulation relation.

\end{itemize}
